DATA 621---Business Analytics and Data Mining

### 2020-09-16 HW1/HW1_draft_kellogg.Rmd
updated selection model

```{r cooks_distance, include=TRUE}
lm_1<- lm(TARGET_WINS ~ TEAM_BATTING_2B +TEAM_BATTING_HR+
          TEAM_BATTING_BB+TEAM_BATTING_SO+ TEAM_BASERUN_SB+TEAM_BASERUN_CS+
          TEAM_PITCHING_HR+ TEAM_PITCHING_H+
          TEAM_FIELDING_E+ TEAM_FIELDING_DP+  SLG, data=train_set)

cooks_dis<- cooks.distance(lm_1)
influential<- as.numeric(names(cooks_dis)[(cooks_dis > (4/nrow(train_set)))])
train_df2<- train_set[-influential, ]

lm_1<- lm(TARGET_WINS ~ TEAM_BATTING_2B +TEAM_BATTING_HR+
          TEAM_BATTING_BB+TEAM_BATTING_SO+ TEAM_BASERUN_SB+TEAM_BASERUN_CS+
          TEAM_PITCHING_HR+ TEAM_PITCHING_H+
          TEAM_FIELDING_E+ TEAM_FIELDING_DP+  SLG, data=train_df2)
knitr::kable(summary(lm_1)$coefficients, digits = 3L, caption = 'Backwards Model Regression Output')
```

@@

```{r model3, include = FALSE}
# Data Transformation & Multicollinearity Diagnostic Measures
train_df.clean = subset(train_set, select = -c(TEAM_PITCHING_HR, TEAM_BATTING_H, TEAM_BATTING_HBP))
stepdata = cbind(train_df.clean,sapply(train_df.clean[2:length(train_df.clean)], 
                                       function(x) x^2))
names(stepdata) = make.unique(names(stepdata), sep = "_")
train.control = trainControl(method = "cv", number = 10)
# Train the model
step.model = train(TARGET_WINS ~ ., data = stepdata, method = "lmStepAIC", 
                   trControl = train.control, trace = FALSE)
# Final model coefficients
step.model$finalModel 
# Summary of the model
mod_sum = summary(step.model$finalModel)
```

@@

In order to select the best model to make predictions, we looked at some measurements that tells us how well each model fits. These include the 1) $R^2$, which represents the proportion of the variance explained by a model; 2) $adj R^2$, which is a modified version of $R^2$; 3) *Root Mean Squared Error* (RMSE), which is the square root of the mean squared error; and 4) *Akaike Information Criterion* (AIC), which is an estimator of out-of-sample prediction error.

It is apparent that Model #2: Pairwise accounts for nearly 43% of the variation in the dependent variable with the independent variables, which is acceptable as a good model ($R^2 = 0.427$, $adjR^2 = 0.417$. Because the RMSE and adjusted $R^2$ statistics already include a minor adjustment for the number of coefficients estimated, to evaluate the model complexity, we compared the AIC. While a smaller AIC is deemed less complex, the difference between Model #2 and Model #3 is quite insignificant.

```{r echo=FALSE}
# Predictions on training data
mod_sum1 = summary(lm_1)  
pred_lm1 = predict(lm_1, newdata = train_set)
pred_step.model = predict(step.model, newdata = stepdata)

# Model accuracy
results = data.frame(Models = c("Model #1: Backwards" , "Model #2: Pairwise", "Model #3: K-fold"),
                R.squared = c(mod_sum1[["r.squared"]], stmr2, mod_sum[["r.squared"]]), 
                adj.R.squared = c(mod_sum[["adj.r.squared"]], stmar2, mod_sum[["adj.r.squared"]]),
                AIC = c(AIC(lm_1), AIC(stepModel), AIC(step.model$finalModel)),
                RSME = c(sqrt(mean((train_set$TARGET_WINS - pred_lm1)^2)), stmRMSE,
                         sqrt(mean((stepdata$TARGET_WINS - pred_step.model)^2))))
results %>% knitr::kable(digits = 3L, caption = 'Performance Statistics & Error Measure of Models' )
```

@@

Using the test data and the selected final model, a comparison in the prediction statistic was conducted.

```{r include=FALSE, eval=FALSE}
# Data Transformation to match train data
eval_set = complete(mice(data = eval_set, method = "pmm", maxit = 5, seed = 525, print = FALSE), 3)

# Add final model
predictions = predict(ADD-MODEL-HERE, newdata = eval_set)

results = cbind(dataset = c('Training Data', 'Test Prediction'), 
                bind_rows(describe(train_set$TARGET_WINS), describe(predictions))) 
rownames(results) = NULL

results[,-c(2,8,11)] %>% knitr::kable(digits = 3L, caption = 'Prediction Comparison with Final Model')
```

@@

>>>>

```{r cooks_distance, include=TRUE}
lm_1<- lm(TARGET_WINS ~ TEAM_BATTING_2B +TEAM_BATTING_HR+
          TEAM_BATTING_BB+TEAM_BATTING_SO+ TEAM_BASERUN_SB+TEAM_BASERUN_CS+
          TEAM_PITCHING_HR+ TEAM_PITCHING_H+
          TEAM_FIELDING_E+ TEAM_FIELDING_DP+  SLG, data=train_set)
cooks_dis<- cooks.distance(lm_1)
influential<- as.numeric(names(cooks_dis)[(cooks_dis > (4/nrow(train_set)))])
train_df2<- train_set[-influential, ]
lm_1<- lm(TARGET_WINS ~ TEAM_BATTING_2B +TEAM_BATTING_HR+
          TEAM_BATTING_BB+TEAM_BATTING_SO+ TEAM_BASERUN_SB+TEAM_BASERUN_CS+
          TEAM_PITCHING_HR+ TEAM_PITCHING_H+
          TEAM_FIELDING_E+ TEAM_FIELDING_DP+  SLG, data=train_df2)

mod_sum1 = summary(lm_1)  
pred_lm1 = predict(lm_1, newdata = train_df2)
knitr::kable(summary(lm_1)$coefficients, digits = 3L, caption = 'Backwards Model Regression Output')
```

@@

```{r model3, include = FALSE}
train_df = read.csv('https://raw.githubusercontent.com/aadler/DT621_Fall2020_Group2/master/HW1/data/moneyball-training-data.csv')
train_df = subset(train_df, select = -c(INDEX, TEAM_BATTING_HBP))
model = lm(TARGET_WINS ~ ., data = train_df)
# Remove Outlier and more cleaning
cooksd = cooks.distance(model)
influential = as.numeric(names(cooksd)[(cooksd > 4*mean(cooksd, na.rm = TRUE))])
train_df = train_df[-c(influential), ]
train_df.clean = complete(mice(data = train_df, method = "pmm", maxit = 5, 
                               seed = 525, print = FALSE), 3)
train_df.clean$TEAM_BATTING_1B = train_df.clean$TEAM_BATTING_H - 
  (train_df.clean$TEAM_BATTING_2B + train_df.clean$TEAM_BATTING_3B + 
     train_df.clean$TEAM_BATTING_HR)
train_df.clean = subset(train_df.clean, select = -c(TEAM_BATTING_H, TEAM_PITCHING_HR))
# Data Transformation & Multicollinearity Diagnostic Measures
stepdata = cbind(train_df.clean,sapply(train_df.clean[2:length(train_df.clean)], 
                                       function(x) x^2))
names(stepdata) = make.unique(names(stepdata), sep = "_")
# Set up repeated k-fold cross-validation
set.seed(525)
train.control = trainControl(method = "cv", number = 10)
# Train the model
step.model = train(TARGET_WINS ~ ., data = stepdata, method = "lmStepAIC", 
                   trControl = train.control, trace = FALSE)
# Final model coefficients
step.model$finalModel 
# Summary of the model
mod_sum = summary(step.model$finalModel)
# Predictions on train data
pred_step.model = predict(step.model, newdata = stepdata)
```

@@

In order to select the best model to make predictions, we looked at some measurements that tells us how well each model fits the training. These include the 1) $R^2$, which represents the proportion of the variance explained by a model; 2) $adj R^2$, which is a modified version of $R^2$; 3) *Root Mean Squared Error* (RMSE), which is the square root of the mean squared error; and 4) *Akaike Information Criterion* (AIC), which is an estimator of out-of-sample prediction error.

When comparing the three models, it was interesting to select which would be our best model given that their performance statistics and error measurements were not significantly different from each other. It is apparent that Model #1: Backwards accounts for nearly 44% of the variation in the dependent variable with the independent variables, which is acceptable as a good model. But it's explanatory power based on an $adj R^2 = 0.414$ suggests it is no different from Model #3: K-fold. In addition, Model #1: Backwards has a highest RSME of all the models, which ranks as one of the major criteria we are using to decide on a model. Now, Model #2: Pairwise resulted in the $R^2 = 0.427$, $adjR^2 = 0.417$, and the lowest $RSME = 10.193$. Because the RMSE and adjusted $R^2$ statistics already include a minor adjustment for the number of coefficients estimated, to evaluate the model complexity, we compared the AIC. While a smaller AIC is deemed less complex, Model #1: Backwards would have been the way to go, however, it was deemed that the AIC differences among the models are quite insignificant. Thus, it is unanimous that Model #2: Pairwise is our final model.

```{r performance, echo=FALSE, message=FALSE, warning=FALSE}
# Model accuracy
results = data.frame(Models = c("Model #1: Backwards" , "Model #2: Pairwise", "Model #3: K-fold"),
                R.squared = c(mod_sum1[["r.squared"]], stmr2, mod_sum[["r.squared"]]), 
                adj.R.squared = c(mod_sum[["adj.r.squared"]], stmar2, mod_sum[["adj.r.squared"]]),
                AIC = c(AIC(lm_1), AIC(stepModel), AIC(step.model$finalModel)),
                RSME = c(sqrt(mean((train_set$TARGET_WINS - pred_lm1)^2)), stmRMSE,
                         sqrt(mean((stepdata$TARGET_WINS - pred_step.model)^2))))
results %>% knitr::kable(digits = 3L, caption = 'Performance Statistics & Error Measure of Models' )
```

@@

Using the test data and the selected final model, Model #2: Pairwise, a comparison in the prediction statistic was conducted.

```{r predictions, echo=FALSE, message=FALSE, warning=FALSE}
# Cleaning test set for Model#1
test_df1<-  eval_set %>% dplyr::select(-c(TEAM_BATTING_HBP, INDEX)) %>%
  mutate(TEAM_BATTING_1B =
         TEAM_BATTING_H -
         (TEAM_BATTING_2B + TEAM_BATTING_3B + TEAM_BATTING_HR)) %>%
  mutate_if(is.integer, as.numeric)
test_df1 <- complete(mice(data = test_df1,
                         method = "pmm",
                         seed = 9450,
                         print=FALSE), 3)

test_df1<-
  test_df1 %>%  mutate(AB = TEAM_BATTING_SO + TEAM_BATTING_BB + TEAM_BATTING_H,
         SLG = (TEAM_BATTING_1B + (2*TEAM_BATTING_2B)+ (3*TEAM_BATTING_3B)+ (4*TEAM_BATTING_HR))/AB) %>% dplyr::select(-c(AB, TEAM_BATTING_H))

# Cleaning test set for Model #2
BxCox <- function(x, l) {
  ifelse(l == 0, log(x), (x ^ l - 1) / l)
}
BCL <- data.table(Metric = c('TEAM_PITCHING_H', 'TEAM_FIELDING_E',
                             'TEAM_BASERUN_SB', 'TEAM_PITCHING_HR',
                             'TEAM_BATTING_HR', 'TEAM_PITCHING_BB',
                             'TEAM_PITCHING_SO', 'TEAM_BATTING_3B',
                             'TEAM_BASERUN_CS'),
                  Lambda = c(-2.5, -0.7, 0, 0.6, 0.6, -0.1, -0.35, 0, 0.5))
setkey(BCL, Metric)
MBT <- setDT(eval_set)
#MBT[, `:=`(TEAM_BATTING_HBP = NULL)]
MBTlng <- melt(MBT, id.vars = 'INDEX', variable.name = 'Metric',
               value.name = 'Value', variable.factor = FALSE)
MBTlng[, Value := as.double(Value)]
MBTlng <- BCL[MBTlng, on = 'Metric']
MBTlng[!is.na(Lambda), Value := BxCox(Value, Lambda)]
MBTlng[, Lambda := NULL]
MBTlngSC <- copy(MBTlng)
# Remove infinities
MBTlngSC[, Value := ifelse(is.infinite(Value), NA, Value)]
# Impute, Center, and Scale
MBTlngSC[, Value := replace(Value, is.na(Value), mean(Value, na.rm = TRUE)),
       keyby = Metric]
MBTlngSC[, Value := scale(Value, center = TRUE, scale = TRUE), keyby = Metric]
MBTSC <- dcast(MBTlngSC, INDEX ~ Metric, value.var = 'Value')

# Cleaning test set for Model #3
test_df = complete(mice(data = eval_set, method = "pmm", maxit = 5, seed = 525, print = FALSE), 3)
test_df$TEAM_BATTING_1B = test_df$TEAM_BATTING_H - 
  (test_df$TEAM_BATTING_2B + test_df$TEAM_BATTING_3B + test_df$TEAM_BATTING_HR)
test_df = cbind(test_df, sapply(test_df[1:length(test_df)], function(x) x^2))
names(test_df) = make.unique(names(test_df), sep = "_")
```

```{r echo=FALSE}
# Model #1
predictions = predict(lm_1, newdata = test_df1)
results = cbind(dataset = c('Training Data', 'Test Prediction'), 
                rbind(psych::describe(train_set$TARGET_WINS), psych::describe(predictions))) 
rownames(results) = NULL
results[,-c(2,8,11)] %>% knitr::kable(digits = 2L, caption = 'Prediction Comparison with Model #1: Backwards')

# Model #2
predictions <- predict(stepModel, MBTSC)
results = cbind(dataset = c('Training Data', 'Test Prediction'), 
                rbind(psych::describe(train_set$TARGET_WINS), psych::describe(predictions))) 
rownames(results) = NULL
results[,-c(2,8,11)] %>% knitr::kable(digits = 2L, caption = 'Prediction Comparison with Model #2: Pairwise')

# Model #3
predictions = predict(step.model$finalModel, newdata = test_df)
results = cbind(dataset = c('Training Data', 'Test Prediction'), 
                rbind(psych::describe(train_set$TARGET_WINS), psych::describe(predictions))) 
rownames(results) = NULL
results[,-c(2,8,11)] %>% knitr::kable(digits = 2L, caption = 'Prediction Comparison with Model #3: K-fold')
```

@@

## CODE APPENDIX

The code chunks below represent the R code called in order during the analysis.
They are reproduced in the appendix for review and comment.
